Compositional Certified Resource Bounds

### 2015-04-07 paper/main.tex
syntax changes

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than $\totalLoc$ lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

@@

\begin{figure}[t]
  \centering
\vspace{-1.5ex}
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K;
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote concrete constants.
    % The reasoning works uniformly for all such $K$ and
    % $T$.
  }
  \label{fig:ex1}
\end{figure}

@@

\begin{figure*}
 \setlength{\progwidth}{.22\linewidth}
  \centering
\vspace{-.3cm}
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y)
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y=y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
} ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

while (x<n) {
  ${\color{blue}%
  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
  if (z>x)
    ${\color{blue}%
    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
  else
    ${\color{blue}%
    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
    z=z+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
  tick(1);
} ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[z, n]|$
\\[.4\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
} ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
   \end{lstlisting}
\vspace{-2.5ex}
$3.1|[y,z]| + 0.1|[0,y]|$
\\[.4\baselineskip]
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
  } ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
} ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
   \end{lstlisting}
\vspace{-2.5ex}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.4\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Derivations of bounds on the number of ticks for challenging
  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
  \emph{t08a} contains \emph{sequential loops} so that the iterations
  of the second loop depend on the first, and \emph{t27} contains
  interacting \emph{nested loops}. \iffull{In the potential functions, we only
  mention the non-zero terms and in the logical context $\Gamma$ we
  only mention assertions that we use.}{} In
  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
  0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}

@@

\ifshort{
\begin{figure*}
\small
\vspace{-.4cm}
\def \MathparLineskip {\lineskip=0.35cm}
\begin{mathpar}
\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
{ }
{ \!\! B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) \!\!}
%
\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\\ \Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\\ \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }
\vspace{-.3cm}
\end{mathpar}
\caption{Selected inference rules of the quantitative analysis.}
\label{fig:auto}
\end{figure*}
}{}

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
%
The part $(\Gamma, Q) \vdash \!\! S \!\! \dashv (\Gamma', Q')$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q(\state')$
\mbox{resources left and $\Gamma'$ holds for $\state'$.}

The judgement is a bit more involved since we have to take into
account the early exit statements \code{break} and \code{return}.
This is similar to classical Hoare triples in the presence of non-linear
control flow. In the judgement, $(\Gamma_B, Q_B)$ is the postcondition
that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
% of looping statements by the rule {\sc Q:Loop} and used before
% \code{break} statements by the rule {\sc Q:Break}.
Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
returning from a function call with \code{return}. \iffull{(This becomes clear
in the rules.)}{}

% These judgements correspond to the logic judgements $\Delta; B; R
% \vdash \htriple P S Q$ where each assertion splits in two parts, a
% logical part $\Gamma$ and a quantitative part $Q$.  That means that
% $(\Gamma_B, Q_B)$ is the postcondition of \code{break} statements,
% $(\Gamma_R, Q_R)$ is the postcondition for return statements, and
% $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ can be understood as a
% simple Hoare triple.  We leave out the function context $\Delta$ in
% the presentation to avoid cluttering notations.  Instead we assume a
% fixed function context $\Delta$ that is implicit on all judgements.
% Function contexts are treated exactly as in the logic.  % We call
% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
% % $(\Gamma', Q')$.
% The correspondence between judgements of the
% automation and the ones of the quantitative logic is made formal by
% our soundness proof of the automatic amortized analysis at the end of
% this section.

@@

During the analysis, function specifications are quadruples
$(\Gamma_f, Q_f, \Gamma'_f, Q'_f)$ where $\Gamma_f, Q_f$ depend on
$\Vargs$, and $\Gamma'_f, Q'_f$ depend on $\Vret$.  These parameters
are instantiated by appropriate variables on call sites.  A distinctive
feature of our analysis is that it respects the function abstraction:
when deriving a function specification it generates a set of
constraints and the above quadruple; once done, the constraint
set can readily be reused for every call site and the function need
not be analyzed multiple times.
%
Therefore, the derivation rules are parametric in a function context $\Delta$
that we leave implicit in the rules presented here. More details
can be found in the extended version.

>>>>

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than $\totalLoc$ lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip

@@

\begin{figure}[t]
  \centering
\vspace{-1.5ex}
\begin{lstlisting}[mathescape]
${\color{blue}%
  \{ \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
while (x+K<=y) {
  ${\color{blue}%
  \{ x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
  x=x+K;
  ${\color{blue}%
  \{ x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
  tick(T);
  ${\color{blue}%
  \{ x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
}
${\color{blue}%
 \{ x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \} }$
\end{lstlisting}
  \caption{Derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote concrete constants.
    % The reasoning works uniformly for all such $K$ and
    % $T$.
  }
  \label{fig:ex1}
\end{figure}

@@

\begin{figure*}
 \setlength{\progwidth}{.22\linewidth}
  \centering
\vspace{-.3cm}
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
  if (m>y)
    ${\color{blue}%
    \{ m{>}y; \, |[x, n]| {+} |[y, m]| \} }$
    y=y+1;
    ${\color{blue}%
    \{ \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  else
    ${\color{blue}%
    \{ n{>}x; \, |[x, n]| {+} |[y, m]| \} }$
    x=x+1;
    ${\color{blue}%
    \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  ${\color{blue}%
  \{ \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \} }$
  tick(1);
} ${\color{blue}%
  \{ \cdot; \, |[x, n]| {+} |[y, m]| \} }$
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

while (x<n) {
  ${\color{blue}%
  \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
  if (z>x)
    ${\color{blue}%
    \{ x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
    x=x+1;
    ${\color{blue}%
    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
  else
    ${\color{blue}%
    \{ z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \} }$
    z=z+1;
    ${\color{blue}%
    \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
  ${\color{blue}%
  \{ \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \} }$
  tick(1);
} ${\color{blue}%
  \{ \cdot; \, |[x, n]| {+} |[z, n]| \} }$
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[z, n]|$
\\[.4\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  \{ y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
  y=y+1;
  ${\color{blue}%
  \{ \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
  tick(3);
  ${\color{blue}%
  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
}
${\color{blue}%
\{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
while (y>9) {
  ${\color{blue}%
  \{ y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
  y=y-10;
  ${\color{blue}%
  \{ \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
  tick(1);
} ${\color{blue}%
  \{ \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \} }$
   \end{lstlisting}
\vspace{-2.5ex}
$3.1|[y,z]| + 0.1|[0,y]|$
\\[.4\baselineskip]
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  \{ n{<}0; \,  P(n,y) \} }$
  n=n+1;
  ${\color{blue}%
  \{ \cdot; \,  59 {+} P(n,y) \} }$
  y=y+1000;
  ${\color{blue}%
  \{ \cdot; \, 9 {+} P(n,y) \} }$
  while (y>=100 && *){
    ${\color{blue}%
    \{ y{>}99; \, 9 {+} P(n,y) \} }$
    y=y-100;
    ${\color{blue}%
    \{ \cdot; \, 14 {+} P(n,y) \} }$
    tick(5);
  } ${\color{blue}%
  \{ \cdot; \, 9 {+} P(n,y) \} }$
  tick(9);
} ${\color{blue}%
  \{ \cdot; \, P(n,y) \} }$
   \end{lstlisting}
\vspace{-2.5ex}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.4\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Derivations of bounds on the number of ticks for challenging
  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
  \emph{t08a} contains \emph{sequential loops} so that the iterations
  of the second loop depend on the first, and \emph{t27} contains
  interacting \emph{nested loops}. \iffull{In the potential functions, we only
  mention the non-zero terms and in the logical context $\Gamma$ we
  only mention assertions that we use.}{} In
  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
  0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}

@@

\ifshort{
\begin{figure*}
\small
\vspace{-.4cm}
\def \MathparLineskip {\lineskip=0.35cm}
\begin{mathpar}
\Rule[leftskip=.4cm,rightskip=.3cm]{Q:Skip}
{ }
{ \!\! B, R \vdash \htriple{ \Gamma; Q }{ \code{skip} }{ \Gamma; Q } \!\!}
%
\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma'; Q'), R \vdash \htriple{ \Gamma; Q }{ S }{ \Gamma;  Q {+} M_l } }
{ B, R \vdash \htriple{ \Gamma; Q }{ \code{loop}~S }{ \Gamma'; Q' } }
%
\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B, R \vdash \htriple{ \Gamma[x/x {+} y]; Q {+} M_u {+} M_e(x {+} y) }{ x \gets x + y }{ \Gamma; Q' }
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B, R \vdash \htriple{ \Gamma[x/x {\pm} y]; Q {+} M }{ x \gets x \pm y }{ \Gamma; Q' }
}
%
\\ \Rule{Q:Seq}
{  B, R \vdash \htriple{ \Gamma; Q }{ S_1 }{ \Gamma'; Q' {+} M_s }
\\\\ B, R \vdash \htriple{ \Gamma'; Q' }{ S_2 }{ \Gamma''; Q'' }
}
{ B, R \vdash \htriple{ \Gamma; Q }{ S_1;S_2 }{ \Gamma''; Q'' } }
%
\and \Rule{Q:Weak}
{ B, R \vdash \htriple{ \Gamma_2; Q_2 }{ S }{ \Gamma'_2; Q'_2 }
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B, R \vdash \htriple{ \Gamma_1; Q_1 }{ S }{ \Gamma'_1; Q'_1 } }
%
\\ \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }
\vspace{-.3cm}
\end{mathpar}
\caption{Selected inference rules of the quantitative analysis.}
\label{fig:auto}
\end{figure*}
}{}

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
$$
%
The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q(\state')$
\mbox{resources left and $\Gamma'$ holds for $\state'$.}

The judgement is a bit more involved since we have to take into
account the early exit statements \code{break} and \code{return}.
This is similar to classical Hoare triples in the presence of non-linear
control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
% of looping statements by the rule {\sc Q:Loop} and used before
% \code{break} statements by the rule {\sc Q:Break}.
Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
returning from a function call with \code{return}. \iffull{(This becomes clear
in the rules.)}{}

% These judgements correspond to the logic judgements $\Delta; B; R
% \vdash \htriple P S Q$ where each assertion splits in two parts, a
% logical part $\Gamma$ and a quantitative part $Q$.  That means that
% $(\Gamma_B; Q_B)$ is the postcondition of \code{break} statements,
% $(\Gamma_R; Q_R)$ is the postcondition for return statements, and
% $(\Gamma; Q) \vdash S \dashv (\Gamma'; Q')$ can be understood as a
% simple Hoare triple.  We leave out the function context $\Delta$ in
% the presentation to avoid cluttering notations.  Instead we assume a
% fixed function context $\Delta$ that is implicit on all judgements.
% Function contexts are treated exactly as in the logic.  % We call
% % precondition the pair $(\Gamma, Q)$ and postcondition the pair
% % $(\Gamma', Q')$.
% The correspondence between judgements of the
% automation and the ones of the quantitative logic is made formal by
% our soundness proof of the automatic amortized analysis at the end of
% this section.

@@

During the analysis, function specifications are quadruples
$(\Gamma_f; Q_f, \Gamma'_f; Q'_f)$ where $\Gamma_f; Q_f$ depend on
$\Vargs$, and $\Gamma'_f; Q'_f$ depend on $\Vret$.  These parameters
are instantiated by appropriate variables on call sites.  A distinctive
feature of our analysis is that it respects the function abstraction:
when deriving a function specification it generates a set of
constraints and the above quadruple; once done, the constraint
set can readily be reused for every call site and the function need
not be analyzed multiple times.
%
Therefore, the derivation rules are parametric in a function context $\Delta$
that we leave implicit in the rules presented here. More details
can be found in the extended version.
